\section{Conclusion}

We presented a dual-encoder segmentation architecture that fuses RGB color features with Canny-derived geometric edge features through multi-scale cross-attention. Three design choices distinguish the proposed model from a naive bottleneck fusion baseline: a learnable scalar gate that allows the model to down-weight noisy edge contributions during training, a second cross-attention module at encoder stage 4 for earlier structural injection, and edge encoder skip connections that propagate boundary cues through the full decoder path. These mechanisms together enable the model to selectively incorporate geometric information while remaining robust to noisy edge inputs.

On the Cataract-Seg dataset, the proposed model achieves an IoU of 0.9530, matching the best-performing baseline (Early Fusion, 0.9532) and outperforming the standard RGB U-Net (0.9529). The results demonstrate that multi-scale cross-attention fusion is competitive with simpler fusion strategies, and that the gated residual formulation prevents performance degradation when the second modality is noisy ,  a practical advantage in clinical settings where edge map quality may vary with imaging conditions. An earlier version of the model without these design improvements ranked last (0.9480 IoU), confirming that each mechanism contributes positively.

The primary limitation of this evaluation is dataset size: with 30 test images, the differences among the top three models are within statistical noise. The dataset also appears to be near saturation for binary segmentation, with all RGB-based models exceeding 0.95 IoU. Future work should validate the proposed architecture on larger ophthalmic datasets and explore learnable edge detectors (e.g., Sobel filters with trainable weights) as a replacement for fixed Canny filtering, which may allow the edge stream to capture more task-relevant structural features. Extension to multi-class segmentation of individual anterior segment structures (cornea, pupil, lens separately) and to intraoperative video frames~\cite{grammatikopoulou2021cadis} are also natural next steps.

Despite these limitations, the proposed model offers a practical and principled approach to multimodal fusion for cataract segmentation. Its lightweight design --- requiring only a single GPU for training and adding modest overhead over a standard U-Net --- makes it suitable for clinical deployment in resource-constrained settings, where large foundation models are not feasible. The use of standard deep learning libraries and the public availability of the source code further lower the barrier for adoption, enabling other researchers to reproduce, validate, and extend the results. The modality-agnostic nature of the architecture further broadens its potential applicability to other medical imaging domains where intra-image multimodal decomposition can provide complementary information.

The gated cross-attention mechanism, multi-scale fusion strategy, and edge encoder skip connections together form a coherent framework for fusing photometric and geometric features. The learnable gate ensures robustness to noisy inputs, the multi-scale attention provides structural context at multiple resolutions, and the edge skips propagate boundary information through the full decoder path. This combination of mechanisms, validated on a real clinical dataset, represents a step toward more reliable and interpretable segmentation models for ophthalmic imaging.
