\section{Introduction}

Cataract disease affects over 100 million people worldwide and remains the leading cause of preventable blindness \cite{wang2024multimodal}. Accurate segmentation of anterior segment structures --- the cornea, pupil, and lens --- from clinical images is essential for automated grading, surgical planning, and intraoperative guidance. However, models that process only the RGB image struggle with two problems: the low color contrast caused by lens opacification (the lens becomes cloudy, making its boundary hard to see in color) and the illumination artifacts common in slit-lamp photography. These issues reduce reliability precisely in the cases that matter most clinically.

A natural way to address this is to give the model access to a second, complementary view of the same image. A \textit{Canny edge map} provides exactly that: it detects boundaries by locating pixels where brightness changes sharply --- these correspond to anatomical edges (cornea boundary, lens capsule) that remain visible even when color contrast is lost. The RGB image carries \textit{photometric} information (color and texture), while the edge map carries \textit{geometric} information (where the boundaries are). Although both come from a single photograph, they represent fundamentally different types of information, which makes their fusion a \textit{multimodal} problem in the architectural sense.

Multimodal fusion architectures that combine complementary representations have proven effective in other medical domains. In brain tumor segmentation, cross-attention between different MRI modalities (T1, T2, FLAIR) consistently outperforms both single-modality and early-concatenation baselines \cite{zhang2022mmformer}. Dual-encoder designs pairing convolutional networks with transformers have shown similar gains \cite{li2024dectnet}. However, these methods assume that multiple modalities come from separate physical sensors. In ophthalmic imaging, obtaining additional modalities (e.g., OCT alongside a slit-lamp photograph) is not always feasible, and no prior work has explored creating an artificial second modality from a single cataract image.

We propose a dual-encoder U-Net that processes the RGB image through one encoder and its Canny edge map through a second, independent encoder. Cross-attention modules at two stages allow the RGB features to ``query'' the edge features for boundary information, enriching the RGB representation with explicit structural cues. A learnable scalar gate controls how much edge information flows in --- it starts turned off (initialized so that $\sigma(\theta) \approx 0.02$) so the model initially behaves like a standard RGB U-Net, and gradually learns to rely on edge cues only where they help. Edge encoder skip connections further propagate geometric information through every level of the decoder.

We evaluate on the Cataract-Seg dataset \cite{cataractseg2023roboflow} against three baselines: a standard RGB U-Net, an edge-only U-Net, and an early-fusion U-Net that concatenates both inputs at the first layer. The proposed model achieves an IoU of 0.9530, matching the strongest baseline, while providing a more principled and interpretable fusion mechanism than simple input concatenation.
